|

### 2020-05-25 paper/Manuscript.Rmd
Update paper

---
title: |
    | WORCS: A Workflow for Open Reproducible Code in the Social Sciences and Beyond <!-- AL: I strongly recommend to make the title more specific. "Science" is very general, and in the Anglo-American understanding actually refers to the beta sciences. Also, the abstract and the paper start with talking about the situation in the social sciences and derive the motivaion from there, so it would be good if the title would reflect this. The acronym can stay as it is. ;) -->
    | `r paste0("Version: ", tryCatch({worcs:::cran_version()}, error=function(err) 'NA'), ", date: ", format(Sys.Date(), "%d %b,\ %Y"))`

@@

abstract: |
  The social sciences are amidst a paradigm shift towards open science. In part,
  this transition has been fueled by cases of scientific fraud and increasing
  awareness of questionable research practices and failed replications. However,
  open science is not
  merely a cure (or punishment) for this crisis - it is also an opportunity.
  Technological advances enable researchers to more easily conduct reliable,
  cumulative, collaborative science. Capitalizing on these advances has the
  potential to accelerate scientific progress. This tutorial paper introduces a
  workflow for open reproducible code in science (WORCS). WORCS is a 
  step-by-step procedure that researchers can follow to make an entire project
  open and reproducible. It is based on best practices, and can be used either
  in parallel to, or in absence of, top-down requirements by journals. WORCS
  lowers the threshold for adopting an open science workflow by providing this
  tutorial, along with an R-package with user-friendly support functions, and an
  RStudio project template.

@@

\vspace{\baselineskip}

<!-- Structure
# Introduction
# Defining open science
* TOP guidelines
# Goal of this paper
* Grass-roots
* open science as opportunity ?move
    - Easier work
    - Less mistakes
    - Better collaboration
    - Cumulative science
    - Error correction
# Introducing the tools
# Setup - do this only once
# WORCS - steps to follow for each project
# Discussion
* advantages
* limitations
* future directions
-->

# Introduction

The (social) sciences are arguably past the tipping point of the paradigm shift
towards open science.
Several immediate causes stirred up support for this transition; including
highly publicized cases of scientific fraud [as discussed by
@leveltFailingScienceFraudulent2012], increasing awareness of questionable
research practices and their consequences for the credibility of findings [@johnMeasuringPrevalenceQuestionable2012], and the replication crisis
[@shroutPsychologyScienceKnowledge2018].
However, open science is not merely a cure for this crisis - it is also an
opportunity [@adolphOpenBehavioralScience2012].
Open science helps researchers to more easily conduct reliable, cumulative, and 
collaborative science [@nosekScientificUtopiaOpening2012].
Capitalizing on these advances has the potential to accellerate scientific
progress [see also @coyneReplicationInitiativesWill2016].

Many individual researchers are motivated to adopt best practices for open
science and enjoy these benefits. And yet, the question of "how" to 
adopt open science practices can be daunting. Making the transition effectively
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by providing a simple workflow 
that meets most requirements open science, based on best practices, and 
supported by user-friendly software in R: The Worfkflow for Open Reproducible 
Code in Science (WORCS). WORCS is a lightweight approach to open science and
computational reproducibility. The workflow (Figure <!--\@ref(fig:workflow)-->) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project open and reproducible. WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them. It can also be used to produce work in accordance with best 
practices in the absence of top-down support. Although the principles underlying
the WORCS procedure are universal, they have been implemented for R users in the
R-package `worcs`, which offers an RStudio project template that is, effectively,
a "one click solution". This paper introduces the WORCS workflow, discusses how
it meets best practices for open science, and illustrates the use of the `worcs`
package.

@@

# What are Open Science practices?

Although Open Science is advocated by many, it does not have a unitary definition.
Instead, several guidelines exist for practices that are considered "open".
The TOP-guidelines are one of the most influential concrete operationalisations
of Open Science principles [@nosekPromotingOpenResearch2015a].
These guidelines describe eight standards for open science: 1) Comprehensive
citation of literature, data, materials, and methods; 2) sharing data, 3)
sharing the code required to reproduce analyses, 4) sharing new research
materials, and 5) sharing details of the design and analysis; 6)
pre-registration of studies before data collection, and 7) pre-registration of
the analysis plan prior to analysis; and 8) replication of published results.
WORCS defines the goals of open science in terms of these guidelines, and is
designed to facilitate meeting each of these guidelines, with one exception: We
do not address replication of published results, because replication relates to
the subject of a research project, not to its execution.
We group the remaining seven guidelines into three categories: Citation (1),
sharing (2-5), and preregistration (6-7).

WORCS meets the TOP-guidelines in a way that is amenable to the 
FAIR Guiding Principles for scientific data management and stewardship [@wilkinsonFAIRGuidingPrinciples2016a]. These principles advocate that research
data should be Findable, Accessible, Interoperable and Reusable. They are
increasingly applied as a standard for other types of research output as well
[@lamprechtFAIRPrinciplesResearch2019].
WORCS helps users meet the FAIR principles by making all research objects -
including data and code - accessible in a repository that can be made public. 
The `worcs` package additionally
facilitates meeting the FAIR principles in several ways. First, each `worcs`
project has a YAML file (https://yaml.org/) that lists the research objects in
the project, which will allow web crawlers 
to index `worcs` projects. Second, a codebook is automatically generated for
each data file stored
using `open_data()` or `closed_data()`. We encourage users to elaborate on the
default codebook by adding
variable description and categories, which might - in the future - enable
indexing data by topic area. 
Incidentally, `worcs` is itself an example of FAIR research software [see @lamprechtFAIRPrinciplesResearch2019 and https://fair-software.eu/]. It is 
hosted on the public version controlled repository 'GitHub', it has an open
source licence (GPL v3.0), is registered in a community registry (CRAN), it
enables the citation of the software (using the `citation("worcs")` command, or
by citing this paper), and followed the CII Best Practices software quality
checklist during development.

@@

Although the principles underlying this workflow are universal, WORCS has 
currently been implemented for R only [@rcoreteamLanguageEnvironmentStatistical2020].
Several arguments support the choice to work in R.
First, R and all of it extensions are free and open source software, which
should be the tool of choice for open science.
Second, as of this writing, R is the most advanced programming language in terms
of the implementation of tools required for an open science workflow.
Third, R is the second-most cited statistical software package [@muenchenPopularityDataScience2012], following SPSS, which has no support for any of the open science tools discussed in this paper.
Fourth, R is well-supported; a vibrant online community exists for R, developing new R methods and packages, and providing support and tutorials for existing ones.
Finally, R is highly interoperable:
Packages are available to load nearly every imaginable filetype, and output can be written to most file types, including DOCX and PDF (in APA style), and HTML format. 
Moreover, wrappers are available for many tools developed in other programming languages, and code written in other programming languages can be evaluated from R [including C++, Fortran, and Python, @allaireMarkdownPythonEngine]. There are excellent free resources for learning R, such as the book R for Data Science [@grolemundDataScience2017].

Working with R is simplified immensely by using the RStudio integrated development engine (IDE) for R [@rstudioteamRStudioIntegratedDevelopment2015].
Most of the tools used in WORCS are embedded directly into the user interface of RStudio.
RStudio automates and streamlines tedious or complicated aspects of working with R, Rmarkdown, and Git/GitHub.
This makes RStudio a comprehensive solution for open science research projects.
Another important feature of RStudio is project management.
A project bundles writing, analyses, data, references, etcetera, into a
self-contained folder, that can be uploaded entirely to 'GitHub' and downloaded
by future users. The `worcs` R-package installs a new RStudio project template.
When a new project is initialized from this template, the bookkeeping required
to set up an open science project is performed automatically. The `worcs` 
package has a [vignette](https://cjvanlissa.github.io/worcs/articles/setup.html)
on how to set up your system with all of the required tools.

@@

WORCS relies on several free, open source software solutions.
The first is *dynamic document generation* (DDG): Writing scientific reports in
a format that interleaves written reports with blocks of code used to conduct
the analyses. The text is automatically formatted; the `worcs` package can 
output manuscripts as an APA-style paper [thanks to the R-package `papaja`, @austPapajaPrepareReproducible2020], or as a host of other scientific formats 
[see @allaireRticlesArticleFormats2020], or as a plain 'Markdown' format for
'GitHub'. When the text is formatted, the results of the code blocks are
automatically generated and insterted in the text, or rendered as Figures and
Tables. Dynamic document generation supercedes
the classical approach of using separate programs to write prose and conduct
analyses, and then manually copy-pasting analysis results into the text.
Although there is a slight learning curve to transitioning to DDG, the
investment really pays off:

* Time saved from painstakingly copy-pasting output and manually formatting text
soon outweighs the investment of switching to a new program
* Human error in manually copying results is eliminated
* When revisions require major changes to the analyses, all results, Figures and
Tables are automatically updated
* Flexibility in output formats means that a manuscript can be rendered to
presentation format, or even to a website
* Reproducibility is guaranteed, and can be verified by reviewers and/or
readers, because the code is run each time the document is compiled.

In sum, while writing academic papers in a programming environment might seem
counter-intuitive at first, this approach is much more amenable to the needs of
academics than most word processing software. It prevents mistakes, and saves
time. We recommend centering a research project around one dynamically generated
(RMarkdown) document, which includes all analysis code. Longer scripts can be
stored in `.R` files, and called from the main document using the `source()`
function.

The second solution is *version control*:
Maintaining an indelible log of every change to all project files. Version 
control is a near-essential tool for scientific reproducibility, as anyone
learns who has had the experience of accidentally deleting a crucial file, or of
being unable to reproduce analyses because they ran some analyses interactively
instead of using syntax [see also @blischakQuickIntroductionVersion2016].
Many scientists use some form of *implicit* version control; for example, by
renaming files after major changes (e.g., "manuscript_final_2.2-2019-10-12.doc"),
tracking changes in word processing software, or using cloud hosting services
that retain backups of previous versions.
WORCS instead uses the *explicit* version control software 'Git'
(www.git-scm.com). Git tracks
changes to files, and stores these changes when the user makes a "commit" (a 
snapshot of the version controlled files).
Git retains a complete log of all commits, and users can compare changes between
different commits, or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
The `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

The functionality of 'Git' is amplified by services such as 'GitHub'
(https://github.com). 'GitHub' is best understood as a cloud storage service
with social networking functionality. The cloud storage aspect
of 'GitHub' works as follows: You can "clone" (copy) a local 'Git' repository to
the 'GitHub' website, as a backup or research archive. The social network aspect
comes into play when a repository on 'GitHub' is set to "public": This allows 
other researchers to peruse the repository and see how the work was done, clone
it to their own computer to replicate the original work or apply the methods to
their own data, open "Issues" to ask questions or give feedback on the project,
or even send a "Pull request" with suggested changes to the text or code for
your consideration. 'Git' and 'GitHub' shine as tools for collaboration, because
different people can simultaneously work on different parts of a project, and
their changes can be compared and automatically merged on the website. Even on
solo projects, working with 'Git'/'GitHub' has many benefits: Staying organized,
being able to start a new study with a clone of an old, similar repository, or
splitting off an "experimental branch" to try something new, while retaining the
ability to "revert" (return) to a previous state of the project, or to "merge"
(incorporate) the experimental branch. This paper focuses on 'GitHub', which is
the most widely used 'Git' cloud storage services. However, as of version 0.1.2,
`worcs` projects can interchangeably use the two main alternative services,
'GitLab' (https://gitlab.com) and 'Bitbucket' (https://bitbucket.org). The
package has a [vignette]() with additional instructions for using these platforms.

The third solution is *dependency management*: Keeping track of exactly what
software was used to conduct the analyses. At first glance, it might seem
sufficient to state that analyses were conducted in *Program X*. However, every
program is susceptible to changes, updates, and bugfixes. R-packages in
particular are updated frequently, because they are Open Source, and there is an
active community of developers contributing functionality and bugfixes. 
Potentially, any update could change the results of the code. <!--For example, 
R-version 3.6.0 fixed a bug in the random number generator. This means that any
analyses that use random numbers, such as bootstrapping, cannot be strictly
replicated from previous versions. Or, for example, R-version 4.0.0 changed a
default parameter that affects how data are loaded from files.-->

<!-- LB: and onwards: this section requires some background knowledge on R/packages/etc. This could be a place where readers stumble and stop reading. Perhaps introduce the concept of the R project template earlier, to see section on dependencies more into perspective. 
[main issue] it should become clear earlier what the WORCS workflow entails: e.g. switching to working with a particular R Project Template. Discuss the ‘workflow' section earlier. Related: Can you make a schematic overview of the WORCS?overview--> 
Many solutions exist to ensure computational reproducibility. These solutions typically work by
enveloping your research project in a distinct "environment" that only has access to programs that
are explicitly installed, and maintaining a record of these programs. These solutions differ in
user-friendliness and effectiveness. If strict reproducibility is required, one might use the
aforementioned Docker-based workflow by @peikertReproducibleDataAnalysis2019, which runs all
analyses inside a "container": An environment that behaves like a virtual computer that can be
stored like a sort of time capsule, and identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the R environment (e.g., preprocessing tools
for neuroimaging data). However, it can be difficult to set up for novice users - although a package
is in development to facilitate this transition [@repro]. The cloud-based research collaboration platform
'Code Ocean' is also based on Docker. It is very user-friendly, but it is not clear whether the
["Data Processing Agreement"](https://www.digitalocean.com/legal/data-processing-agreement/) is fully
GDPR compliant - it seems to focus more on limiting liability than on compliance. This is a risk with
any cloud-computing solution.

For analyses conducted entirely in R, a slightly less sophisticated solution is offered by the R-package `packrat`, which installs all packages
used in a project to a dedicated library that can be stored and copied with the project. This has the disadvantage that for each project, all packages
must be installed anew, which can take a long time (sometimes hours) and a lot of hard drive space.
Moreover, 'GitHub' is not designed to store your entire project library, so `packrat` does not address
the issue of how other users are supposed to get access to your library. In conclusion, `packrat` is
somewhat unwieldy. The package `checkpoint` offers an extremely lightweight solution: Instead of storing
all R-packages used, it simply records the calendar date, and when another user loads your project, it installs the version of all required
packages that was available on the central R repository `CRAN` on that date from an archival website.
Thus, all that is needed to restore a project is the date on which it was conducted. The main limitation is that this solution is limited to `CRAN`
packages, and cannot mix package versions from different dates.

A recently developed package strikes a great balance between user-friendliness 
and effectiveness: `renv`, developed by the team behind 'RStudio'. This package
maintains a text-based, human-readable log of the packages used, their version
numbers, and where they were installed from (e.g., 'CRAN', 'Bioconductor', 'GitHub').
This text-based log can be version controlled with 'Git'. `renv` installs all
packages in a cache that is shared between all projects that use `renv`. Thus,
if one of your projects uses a package that is already in the cache, it does not
need to be installed again. This overcomes the limitations of long installation
times and large space requirements of `packrat`. When someone else loads your 
project, `renv` will install all of the required packages onto a cache on their
computer. Whereas `packrat` copies the source code of all packages, `renv`
obtains the packages from their original repositories. Thus, as long as these
repositories are maintained, this solution is very lightweight and effective.
It is not foolproof, but all solutions for computational
reproducibility have a limited shelf life [@brownHowLearnedStop2017].
Ultimately, the best solution might be to use a good-enough solution, like
`renv`, and acknowledge that all code requires some maintenance if you want to
reproduce it in the future.

@@

A key consideration when developing a research project is what filetypes to use.
A case can be made to use text-based files only (or as much as possible),
instead of binary files.
Text-based files can be read by machines and humans alike.
Binary files, such as 'Word' (`.docx`) or 'SPSS' (`.sav`, `.spo`) files,
must be decoded by (commercial) software first.
'Git' is designed to track any changes to the content of text-based files.
Each change reflects a line of text added or removed. This means that, when
writing a version-controlled paper in plain-text, it is recommended to insert a
line break after every sentence (instead of after every paragraph). That way,
the change log will indicate which specific sentence was edited.
Binary files should not be version controlled with 'Git',
as changes in the encrypted file are uninterpretable. Binary files are often 
also larger than text-based files, which means they take up more space on cloud
hosting services, such as 'GitHub'. For those reasons, uploading large binary
files to these platforms is frowned upon.
'GitHub' does allow visitors to peruse text-based files online, and
conveniently renders certain filetypes: For example, 'GitHub' displays `.md`
(Markdown) files as web pages, and `.csv` files as spreadsheets. This is why
`worcs` uses Markdown and `.csv` extensively.

@@

Before you can use the WORCS workflow, you have to install the required software. You have to execute this setup procedure only once for each computer you intend to use `worcs` on. The procedure should take about 30 minutes, and it is documented in [the setup vignette](https://cjvanlissa.github.io/worcs/articles/setup.html). After preparing your system, you can use the WORCS workflow for all of your projects. This step-by-step procedure is detailed in [the workflow vignette](https://cjvanlissa.github.io/worcs/articles/workflow.html). In this paper, we provide only a brief outline of the WORCS procedure (as of Version `r cran_version()`) below. We highly recommend users to refer to [the workflow vignette](https://cjvanlissa.github.io/worcs/articles/setup.html) for the most up-to-date procedure (subject to improvements).

@@

TOP encourages comprehensive citation of literature, data, materials, methods,
and software. In principle,
researchers can meet this requirement by simply
citing every reference used. Unfortunately, citation of data and software is
less commonplace than citation of literature and materials. Crediting these
resources is important, because it incentivizes data sharing and the development
of open-source software, supports the open science efforts of others, and helps
researchers receive credit for *all* of their research output, in line with
[DORA](https://sfdora.org/).

@@

When writing a manuscript in Rmarkdown, analysis code is embedded in the prose of the paper. Thus, the TOP-guideline of sharing analysis code can be met simply by committing the source code of the manuscript to GitHub, and making this GitHub repository Public upon publication. If authors additionally use open data and a reproducible environment (default in WORCS), then a third party can simply replicate all analyses by copying the entire GitHub repository, and Knitting the manuscript on their local computer.

Aside from analysis code, the TOP-guidelines also encourage sharing new research materials, and details of the study design and analysis. These goals can be accomplished simply by placing any relevant documents in the RStudio project folder, committing them to the Git repository, and pushing to GitHub. As with any document version controlled in this way, it is advisable to use plain text only.

>>>>

---
title: |
    | WORCS: A Workflow for Open Reproducible Code in Science
    | `r paste0("Version: ", worcs:::cran_version(), ", date: ", format(Sys.Date(), "%d %b,\ %Y"))`

@@

abstract: |
  Technological advances enable researchers to more easily conduct reliable,
  cumulative, collaborative science. Capitalizing on these advances has the
  potential to accelerate scientific progress. However, adopting open science
  principles can be challenging and costly, because doing so requires
  substantial conceptual education and training in the use of new tools. The 
  goal of this paper is to remove these obstacles. We introduce the Workflow for
  Open Reproducible Code in Science (WORCS). WORCS constitutes a step-by-step
  procedure that researchers can follow to make an entire project open and
  reproducible. It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies. To further facilitate the transition, the
  WORCS principles are implemented in an R-package, which offers an RStudio 
  project template that is, effectively, a "one-click solution" for R users. 
  This paper introduces the workflow, discusses how it meets different 
  requirements of open science, and provides examples of the implementation of
  `worcs` in R.

@@

<!--\vspace{\baselineskip}-->

The (social) sciences are arguably past the tipping point of a paradigm shift
towards open science.
Several immediate causes rallied support for this transition, including
highly publicized cases of scientific fraud [as discussed by
@leveltFailingScienceFraudulent2012], increasing awareness of questionable
research practices and their consequences for the credibility of findings [@johnMeasuringPrevalenceQuestionable2012], and the replication crisis
[@shroutPsychologyScienceKnowledge2018].
However, open science should not be seen as a cure (or punishment) for this
crisis. As the late dr. Jonathan Tennant put it: "Open science is just good science"
[@tennantOpenScienceJust2018].
Moreover, open science creates opportunities for researchers to more easily
conduct reliable, cumulative, and 
collaborative science [@nosekScientificUtopiaOpening2012;
@adolphOpenBehavioralScience2012]. Open science also
promotes inclusivity, because it removes barriers for participation.
Capitalizing on these advances has the potential to accellerate scientific
progress [see also @coyneReplicationInitiativesWill2016].

Many individual researchers are motivated to adopt best practices for open
science and enjoy these benefits. And yet, the question of "how" to 
adopt open science practices can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by providing a simple workflow 
that meets most requirements open science, is based on best practices, and 
is supported by user-friendly software in R: The Worfkflow for Open Reproducible 
Code in Science (WORCS). WORCS is a lightweight approach to open science and
computational reproducibility. The workflow (Figure <!--\@ref(fig:workflow)-->) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project open and reproducible. WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them. It can also be used in the absence of top-down support or guidelines
to produce work in accordance with best practices.
Although the principles underlying
the WORCS procedure are universal, they have been implemented for R users in the
R-package `worcs`, which offers an RStudio project template that is, effectively,
a "one click solution". This paper introduces the WORCS workflow, discusses how
it meets best practices for open science, and illustrates the use of the `worcs`
package.

@@

# Defining Open Science practices

Although Open Science is advocated by many, it does not have a unitary definition
[but, for an effort to ground open science in first principles, see 
@tennantValuePropositionOpen2020].
Instead, several guidelines exist for practices that are considered "open".
The TOP-guidelines are one of the most influential concrete operationalisations
of Open Science principles [@nosekPromotingOpenResearch2015a].
These guidelines describe eight standards for open science: 1) Comprehensive
citation of literature, data, materials, and methods; 2) sharing data, 3)
sharing the code required to reproduce analyses, 4) sharing new research
materials, and 5) sharing details of the design and analysis; 6)
pre-registration of studies before data collection, and 7) pre-registration of
the analysis plan prior to analysis; and 8) replication of published results.
WORCS defines the goals of open science in terms of these guidelines, and is
designed to facilitate meeting each of these guidelines, with one exception: We
do not address replication of published results, because replication relates to
the subject of a research project, not to its execution.
We group the remaining seven guidelines into three categories: citation (1),
sharing (2-5), and preregistration (6-7).

@@

Although the principles underlying this workflow are universal, WORCS has first
been implemented for R users [@rcoreteamLanguageEnvironmentStatistical2020].
Several arguments support the choice to work in R.
First, R and all of it extensions are free open source software, which
should be the tool of choice for open science.
Second, as of this writing, R is the most advanced programming language in terms
of the implementation of tools required for an open science workflow.
Third, R is the second-most cited statistical software package [@muenchenPopularityDataScience2012], following SPSS, which has no support for any of the open science tools discussed in this paper.
Fourth, R is well-supported; a vibrant online community exists for R, developing new R methods and packages, and providing support and tutorials for existing ones.
Finally, R is highly interoperable:
Packages are available to load nearly every imaginable filetype, and output can be written to most file types, including DOCX and PDF (in APA style), and HTML format. 
Moreover, wrappers are available for many tools developed in other programming languages, and code written in other programming languages can be evaluated from R, including C++, 
Fortran, and Python [@allaireMarkdownPythonEngine]. There are excellent free
resources for learning R, such as the book R for Data Science [@grolemundDataScience2017].

Working with R is simplified immensely by using the RStudio integrated
development engine (IDE) for R [@rstudioteamRStudioIntegratedDevelopment2015].
RStudio automates and streamlines tedious or complicated aspects of working with
the tools used in WORCS, and many are embedded directly into the visual user
interface.
This makes RStudio a comprehensive solution for open science research projects.
Another important feature of RStudio is project management.
A project bundles writing, analyses, data, references, etcetera, into a
self-contained folder, that can be uploaded entirely to 'GitHub' and downloaded
by future users. The `worcs` R-package installs a new RStudio project template.
When a new project is initialized from this template, the bookkeeping required
to set up an open science project is performed automatically. The `worcs` 
package has a [vignette](https://cjvanlissa.github.io/worcs/articles/setup.html)
on how to set up your system with all of the required tools.

@@

### Dynamic document generation

WORCS relies on several free, open source software solutions.
The first is *dynamic document generation* (DDG): Writing scientific reports in
a format that interleaves written reports with blocks of code used to conduct
the analyses. The text is automatically formatted as a scientific paper in 
several potential styles. When the text is formatted, the results of the code
blocks are
automatically generated and insterted in the text, or rendered as Figures and
Tables. Dynamic document generation supercedes
the classical approach of using separate programs to write prose and conduct
analyses, and then manually copy-pasting analysis results into the text.

Although there is a slight learning curve to transitioning to DDG, the
investment really pays off. Time saved from painstakingly copy-pasting output
and manually formatting text soon outweighs the investment of switching to a new
program. Moreover, human error in manually copying results is eliminated.
When revisions require major changes to the analyses, all results, Figures and
Tables can be automatically updated. The flexibility in output formats also 
means that a manuscript can be rendered to presentation format, or even to a
website or blog post. Moreover, reproducibility is guaranteed, and can be
verified by reviewers and/or readers, because the code is run each time the
document is compiled. In sum, while writing academic papers in a programming environment might seem
counter-intuitive at first, this approach is much more amenable to the needs of
academics than most word processing software. It prevents mistakes, and saves
time.

In the R-implementation of WORCS, we recommend centering a research project
around one dynamically generated 'RMarkdown' document, which includes all 
analysis code. Longer scripts can be
stored in `.R` files, and called from the main document using the `source()`
function. When a reader or reviewer compiles this document, all code is run
automatically, thus verifying computational replicability. The 'RMarkdown' 
document can be automatically formatted in many styles, including
APA style [thanks to the R-package `papaja`, @austPapajaPrepareReproducible2020],
a host of other scientific formats 
[see @allaireRticlesArticleFormats2020], and as plain 'Markdown' for 'GitHub'.

### Version control

The second solution is *version control*:
Maintaining an indelible log of every change to all project files. Version 
control is a near-essential tool for scientific reproducibility, as anyone
learns who has had the experience of accidentally deleting a crucial file, or of
being unable to reproduce analyses because they ran some analyses interactively
instead of using syntax [see also @blischakQuickIntroductionVersion2016].
Many scientists use some form of *implicit* version control; for example, by
renaming files after major changes (e.g., "manuscript_final_2.2-2019-10-12.doc"),
tracking changes in word processing software, or using cloud hosting services
that retain backups of previous versions.

WORCS recommends using the *explicit* version control software 'Git'
(www.git-scm.com). Git tracks
changes to files, and stores these changes when the user makes a "commit" (a 
snapshot of the version controlled files).
Git retains a complete log of all commits, and users can compare changes between
different commits, or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
A `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

The functionality of 'Git' is amplified by services such as 'GitHub'
(https://github.com). 'GitHub' is best understood as a cloud storage service
with social networking functionality. The cloud storage aspect
of 'GitHub' works as follows: You can "clone" (copy) a local 'Git' repository to
the 'GitHub' website, as a backup or research archive. The social network aspect
comes into play when a repository on 'GitHub' is set to "public": This allows 
other researchers to peruse the repository and see how the work was done, clone
it to their own computer to replicate the original work or apply the methods to
their own data, open "Issues" to ask questions or give feedback on the project,
or even send a "Pull request" with suggested changes to the text or code for
your consideration. 'Git' and 'GitHub' shine as tools for collaboration, because
different people can simultaneously work on different parts of a project, and
their changes can be compared and automatically merged on the website. Even on
solo projects, working with 'Git'/'GitHub' has many benefits: Staying organized,
being able to start a new study with a clone of an old, similar repository, or
splitting off an "experimental branch" to try something new, while retaining the
ability to "revert" (return) to a previous state of the project, or to "merge"
(incorporate) the experimental branch.

The R-implementation of WORCS uses the package `gert`, a 'Git' client for 'R'
[@oomsGertSimpleGit2019], to connect a local project to a remote repository on
'GitHub'. The `worcs` package also contains a user-friendly function to set
the default user credentials, `git_user()`, and a function to add, commit, and
push all changed files in one step, `git_update()`. It is worth noting that,
although this paper focuses on 'GitHub', which is the most widely used 'Git'
cloud storage service, two notable alternative services exist: 
'GitLab' (https://gitlab.com) and 'Bitbucket' (https://bitbucket.org). As of
version 0.1.2, the `worcs` template dialog accepts 'https' addresses of 
repositories on both of these platforms interchangeably. The package has a
[vignette](https://cjvanlissa.github.io/worcs/articles/git_cloud.html) with
additional instructions for using these platforms. However,
`worcs` has been most extensively tested with 'GitHub'.

### Dependency management

The third solution is *dependency management*: Keeping track of exactly what
software was used to conduct the analyses. At first glance, it might seem
sufficient to state that analyses were conducted in *Program X*. However, every
program is susceptible to changes, updates, and bugfixes. Open Source software,
in particular, is regularly updated because there is an
active community of developers contributing functionality and bugfixes. 
Potentially, any such update could change the results of the code, thus rendering
the analysis 'computationally non-reproducible'.

<!-- LB: and onwards: this section requires some background knowledge on R/packages/etc. This could be a place where readers stumble and stop reading. Perhaps introduce the concept of the R project template earlier, to see section on dependencies more into perspective. 
[main issue] it should become clear earlier what the WORCS workflow entails: e.g. switching to working with a particular R Project Template. Discuss the ‘workflow' section earlier. Related: Can you make a schematic overview of the WORCS?overview--> 
Many solutions exist to ensure computational reproducibility. These solutions typically work by
enveloping your research project in a distinct "environment" that only has access to programs that
are explicitly installed, and maintaining a record of these programs. These solutions differ in
user-friendliness and effectiveness. If strict reproducibility is required, users might choose to conduct all analyses in a 'Docker container': An environment that behaves like a
virtual computer, which can be stored like a sort of time capsule, and
identically reinstated on a user computer, or in the cloud.
This is effective, and also preserves software outside the R environment (e.g.,
preprocessing tools for neuroimaging data). An excellent Docker-based workflow
was recently introduced by @peikertReproducibleDataAnalysis2019. This workflow
is based on many of the same principles as WORCS, and the two approaches are
conceptually and practically compatible. However, the workflow described by 
Peikert and Brandmaier can be difficult to set up for novice users. An R-package
is currently in development to facilitate this transition [@repro]. Another 
user-friendly Docker-based workflow is offered by the cloud-based collaboration
platform 'Digital Ocean'. Although Digital Ocean enjoys some popularity among 
scientists, it is not clear whether this platform's
["Data Processing Agreement"](https://www.digitalocean.com/legal/data-processing-agreement/)
is GDPR compliant - it seems to focus more on limiting liability than on
compliance. The necessity to upload data for analysis is a risk with any
cloud-computing solution when human participants are involved.

For analyses conducted entirely in R, the recently released package `renv`
strikes a great balance between user-friendliness  and effectiveness. Developed by
the team behind 'RStudio', this package simply maintains a text-based,
human-readable log of all packages used, their version numbers, and where they
were installed from (e.g., 'CRAN', 'Bioconductor', 'GitHub'). As long as these
repositories maintain their archives, this solution is reliable and very
lightweight.
The text-based log file can be version controlled with 'Git'. `renv`
installs the logged packages in a cache that is shared between all
projects on the computer. Thus, if one of your projects uses a package that is
already in the cache, it does not
need to be installed again. This prevents long installation
times, large space requirements, and frustration having to re-install software
that is already installed on the system. When someone else loads your 
project, `renv` will install all of the required packages onto a cache on their
computer. 
This is not a foolproof method to ensure strict computational reproducibility, but it is more user-friendly
and lightweight than any alternative out there, requiring no technical skill to
setup. When choosing an appropriate solution for dependency management, it is
important to consider that all solutions have a limited shelf life
[@brownHowLearnedStop2017].
Ultimately, the best approach might be to use a "just good-enough" solution, and
acknowledge that all code requires some maintenance if you want to
reproduce it in the future. For projects in R, `renv` meets these requirements.

@@

A key consideration when developing a research project is what filetypes to use.
A case can be made to use text-based files exclusively (or as much as possible),
instead of binary files.
Text-based files can be read by machines and humans alike.
Binary files, such as 'Word' (`.docx`) or 'SPSS' (`.sav`, `.spo`) files,
must be decoded first, and many require commercial software to do so.
Text-based files are also required when using 'Git' for version control.
In 'Git', each change reflects a line of text added or removed. This means that, when
writing a version-controlled paper in plain-text, it is recommended to insert a
line break after every sentence (instead of after every paragraph). That way,
the change log will indicate which specific sentence was edited.
Binary files should not be version controlled with 'Git',
as changes in the encrypted file are uninterpretable. Binary files are also often 
larger than text-based files, which means they take up more space on cloud
hosting services, such as 'GitHub'. For these reasons, uploading large binary
files to these platforms is frowned upon.
'GitHub' encourages the use of text-based files by allowing visitors of their website to peruse
text-based files online, and
conveniently renders certain filetypes: For example, 'GitHub' displays `.md`
(Markdown) files as web pages, and `.csv` files as spreadsheets. This is why
`worcs` uses text-based file types like Markdown and `.csv` extensively.

@@

Before you can use the R-implementation of the WORCS workflow, you have to install the required software.
You only have to perform this setup procedure only once for each computer you
intend to use `worcs` on. It should take about 30 minutes, and it is documented
in [the setup vignette](https://cjvanlissa.github.io/worcs/articles/setup.html).
The setup vignette does not assume a prior installation of R, so it is suitable for novice users.
After setting up your system, you can use the WORCS workflow for all of your
projects. This step-by-step procedure is detailed in [the workflow vignette](https://cjvanlissa.github.io/worcs/articles/workflow.html).
In this paper, we provide only a brief outline of the WORCS procedure (as of Version `r worcs:::cran_version()`) below.
We highly recommend users to refer to [the workflow vignette](https://cjvanlissa.github.io/worcs/articles/setup.html) for the most up-to-date procedure (subject to improvements).

@@

The TOP-guidelines encourage comprehensive citation of literature, data, materials, methods,
and software. In principle,
researchers can meet this requirement by simply
citing every reference used. Unfortunately, citation of data and software is
less commonplace than citation of literature and materials. Crediting these
resources is important, because it incentivizes data sharing and the development
of open-source software, supports the open science efforts of others, and helps
researchers receive credit for *all* of their research output, in line with
[DORA](https://sfdora.org/).

@@

When writing a manuscript created using dynamic document generation, analysis
code is embedded in the prose of the paper. Thus, the TOP-guideline of sharing
analysis code can be met simply by committing the source code of the manuscript
to the 'Git' repository, and making this repository Public, e.g., on 'GitHub'.
If authors additionally use open data and a reproducible environment (as suggested
in WORCS), then a third party can simply replicate all analyses by copying the
entire repository from the 'Git' hosting service, and Knitting the manuscript on
their local computer.

Aside from analysis code, the TOP-guidelines also encourage sharing new research
materials, and details of the study design and analysis. These goals can be
accomplished by placing any such documents in the 'Git' repository folder,
committing them, and pushing to a cloud hosting service. As with any document
version controlled in this way, it is advisable to use plain text only.
